Program Transformations for miniKanren

### 2020-05-23 papers/miniKanren/text/nonconj.tex
Nonconj done

Hereafter, we consider all goals and relation bodies to be in \emph{canonical normal form}~---~a disjunction of conjunctions of either calls or unifications.
Moreover, we assume all fresh variables to be introduced into scope and all unifications to be computed at each step.
Those disjuncts in which unifications fail are removed.
Other disjuncts take form of a conjunction of relation calls accompanied with a substitution computed from unificaitons.
Any miniKanren term can be trivially transformed to the described form.
In Fig.~\ref{fig:ncpd-pseudo} function \code{normalize} is assumed to perform term normalization.
The code is omitted for brevity.

@@

There are several core ideas behind this algorithm.
The first is to select an arbitrary relation to unfold, not necessarily the leftmost which is safe.
The second idea is to use a heuristic which decides if unfolding a relation call can lead to discovery of contradictions between conjuncts which in turn leads to restriction of the answer set at specialization-time \db{(line 14; \code{heuristically\_select\_a\_call} stands for heurictics combination, see section~\ref{sec:heurictic} for details)}.
If those contradictions are found, then they are exposed by considering the conjunction as a whole and substituting the result of the call unfolding back into the conjunction the call was selected from thus \emph{joining} the conjunction back together instead of using \emph{split} as in CPD \db{(lines 15--22)}.
Finally, if the heuristic fails to select a potentially good call, then the conjunction is splitted into individual calls which are driven in isolation and are never joined \db{(line 23)}.

When the heuristic selects a call to unfold (line \db{15}), a process tree is constructed for the selected call \emph{in isolation} (line \db{16}).
The leaves of the computed tree are examined.
If all leaves are either computed substitutions or are renamings of some relations accompanied with non-empty substitutions, then the leaves are collected and replace the considered call in the root conjunction (lines \db{19--20}).
If the selected call does not suit the criteria, the results of its unfolding is not propagated onto other relation calls withing the conjunction and the next suitable call is selected (line \db{22}).
According to the denotational semantics of \mk{} it is safe to compute individual conjuncts in any order, thus it is ok to drive any call and then propagate its results onto the other calls.

@@

This process creates branchings whenever a disjunction is examined lines 4--\db{5}).
At each step we make sure that we do not start driving a conjunction which we have already examined.
To do this, we check if the current conjunction is renaming of any other configuration in the tree (line \db{11}).
If it is, then we create a special node which then is residualized into a call to the corresponding relation.
This is known as \emph{folding} in supercompilation.

We decided not to perform generalization in this approach in the same fashion as CPD or supercompilation does.
Our conjunctions are always splitted into individual calls and are joined back together if it is sensible.
If the need for generalization arises, i.e. embedding is detected, then we immediately stop driving this conjunction (line \db{12}).
When redisualizatin such conjunction, we just generate a conjunction of calls to the input program before specialization.

@@

Unfolding in our case is done by substitution of some relation call by its body with simultaneous normalization and computation of unifications.
% To unfold a relation call we do the following steps.
% \db{TODO: the following has alredy mentioned above.}
% First, the formal arguments of a relation are substituted for the actual arguments of the call in the body.
% All fresh variables get instantiated.
% The body is transformed into a canonical form (disjunction of conjunctions of either calls or unifications).
% All unifications are computed.
% Those disjuncts in which unifications fails are removed.
% Other disjuncts take form of a conjunction of relation calls accompanied with a substitution.
The unfolding itself is straightforward however it is not always clear what to unfold and when to \emph{stop} unfolding.
Unfolding in functional programming languages specialization, as well as inlining in imperative one, is usually considered to be safe from a residual program efficiency point of view.
It may only lead to code explosion or code duplication which is mostly left to a target program compiler optimization or even is out of consideration at all if a specializer is considered as a standalone tool~\cite{jonesbook}.

@@

There is a fine edge between too much unfolding and not enough unfolding.
The former may be even worse than the latter.
We believe that the following heuristic provides a reasonable approach to controling unfolding.

@@

We deem every static conjunct (non-recursive) to be safe because they never lead to growth in the number of conjunctions.
Those calls which unfold deterministically, meaning there is only one disjunct in the unfolded relation, are also considered to be safe.

The other part of heuristic is less trivial, we call it the branching heuristic.
It considers the case when the unfolded relation contains less disjuncts than it could possibly have.
This means that we found some contradiction, some computations were gotten rid of, and thus the answer set was restricted, which is desirable when unfolding.
To compute this heuristic we precompute the maximum possible number of disjuncts in each relation and compare this number with the number of disjuncts when unfolding a concrete relation call.
Maximum number of disjuncts is computed by unfolding the body of the relation in which all relation calls were replaced by a unification which always succeeds.

@@

The pseudocode describing heuristic is shown in fig.~\ref{fig:heu-pseudo}.
Selecting a good relation call can fail (line 1).
The implementation works such that we first select those relation calls which are static, and only if there are none, we proceed to look at deterministic unfoldings and then we search for those which are less branching.
This heuristic provides a good balance in unfolding.
% The final heuristic selects the first conjunct which suites either of the following cases.
% First we unfold those conjuncts which are static.
% Then --- deterministic.
% Then those which are less branching.
% The last to be unfolded are those calls, which unfold to a substitution with not conjunction.

>>>>

Hereafter, we consider all goals and relation bodies to be in \emph{canonical normal form}~---~a disjunction of conjunctions of either calls or unifications.
Moreover, we assume all fresh variables to be introduced into scope and all unifications to be computed at each step.
Those disjuncts in which unifications fail are removed.
Other disjuncts take form of a possibly empty conjunction of relation calls accompanied with a substitution computed from unificaitons.
Any \mk{} term can be trivially transformed to the described form.
In Fig.~\ref{fig:ncpd-pseudo} function \code{normalize} is assumed to perform term normalization.
The code is omitted for brevity.

@@

There are several core ideas behind this algorithm.
The first is to select an arbitrary relation to unfold, not necessarily the leftmost which is safe.
The second idea is to use a heuristic which decides if unfolding a relation call can lead to discovery of contradictions between conjuncts which in turn leads to restriction of the answer set at specialization-time (line 14; \code{heuristically\_select\_a\_call} stands for heurictics combination, see section~\ref{sec:heurictic} for details).
If those contradictions are found, then they are exposed by considering the conjunction as a whole and replacing the selected relation call with the result of its unfolding thus \emph{joining} the conjunction back together instead of using \emph{split} as in CPD (lines 15--22).
Joining instead of splitting is why we call our transformer \emph{non-conjunctive} partial deduction.
Finally, if the heuristic fails to select a potentially good call, then the conjunction is splitted into individual calls which are driven in isolation and are never joined (line 23).

When the heuristic selects a call to unfold (line 15), a process tree is constructed for the selected call \emph{in isolation} (line 16).
The leaves of the computed tree are examined.
If all leaves are either computed substitutions or are instances of some relations accompanied with non-empty substitutions, then the leaves are collected and each of them replace the considered call in the root conjunction (lines 19--20).
If the selected call does not suit the criteria, the results of its unfolding is not propagated onto other relation calls within the conjunction, instead the next suitable call is selected (line 22).
According to the denotational semantics of \mk{} it is safe to compute individual conjuncts in any order, thus it is ok to drive any call and then propagate its results onto the other calls.

@@

This process creates branchings whenever a disjunction is examined (lines 4--5).
At each step we make sure that we do not start driving a conjunction which we have already examined.
To do this, we check if the current conjunction is a renaming of any other configuration in the tree (line 11).
If it is, then we fold the tree by creating a special node which then is residualized into a call to the corresponding relation.

We decided not to perform generalization in this approach in the same fashion as CPD or supercompilation does.
Our conjunctions are always splitted into individual calls and are joined back together only if it is sensible.
If the need for generalization arises, i.e. homeomorphic embedding of conjunctions~\cite{de1999conjunctive} is detected, then we immediately stop driving this conjunction (line 12).
When redisualizating such conjunction, we just generate a conjunction of calls to the input program before specialization.

@@

Unfolding in our case is done by substitution of some relation call by its body with simultaneous normalization and computation of unifications.
% To unfold a relation call we do the following steps.
% \db{TODO: the following has alredy mentioned above.}
% First, the formal arguments of a relation are substituted for the actual arguments of the call in the body.
% All fresh variables get instantiated.
% The body is transformed into a canonical form (disjunction of conjunctions of either calls or unifications).
% All unifications are computed.
% Those disjuncts in which unifications fails are removed.
% Other disjuncts take form of a conjunction of relation calls accompanied with a substitution.
The unfolding itself is straightforward however it is not always clear what to unfold and when to \emph{stop} unfolding.
Unfolding in functional programming languages specialization, as well as inlining in imperative one, is usually considered to be safe from the point of view of residual program efficiency.
It may only lead to code explosion or code duplication which is mostly left to a target program compiler optimization or even is out of consideration at all if a specializer is considered as a standalone tool~\cite{jonesbook}.

@@

There is a fine edge between too much unfolding and not enough unfolding.
The former may be even worse than the latter.
We believe that the following heuristic provides a reasonable approach to unfolding control.

@@

We deem every static (non-recursive) conjunct to be safe because they never lead to growth in the number of conjunctions.
Those calls which unfold deterministically, meaning there is only one disjunct in the unfolded relation, are also considered to be safe.

Those relation calls which are neither static nor deterministic are examined with what we call the \emph{less-branching} heuristic.
It identifies the case when the unfolded relation contains less disjuncts than it could possibly have.
This means that we found some contradiction, some computations were gotten rid of, and thus the answer set was restricted, which is desirable when unfolding.
To compute this heuristic we precompute the maximum possible number of disjuncts in each relation and compare this number with the number of disjuncts when unfolding a concrete relation call.
Maximum number of disjuncts is computed by unfolding the body of the relation in which all relation calls were replaced by a unification which always succeeds.

@@

The pseudocode describing heuristic is shown in fig.~\ref{fig:heu-pseudo}.
Selecting a good relation call can fail (line 1).
The implementation works such that we first select those relation calls which are static, and only if there are none, we proceed to consider deterministic unfoldings and then we search for those which are less branching.
We believe this heuristic provides a good balance in unfolding.
% The final heuristic selects the first conjunct which suites either of the following cases.
% First we unfold those conjuncts which are static.
% Then --- deterministic.
% Then those which are less branching.
% The last to be unfolded are those calls, which unfold to a substitution with not conjunction.
